dyngen: benchmarking with in silico single cells

### 2019-12-09 manuscript.Rmd
update

---
title: "dyngen: benchmarking with in silico single cells"
author:
- Robrecht Cannoodt
- Wouter Saelens
- Louise Deconinck
- Yvan Saeys
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    citation_package: biblatex
biblatexoptions: sorting=none
classoption:
- table
- 10pt
- a4paper
header-includes: |
  \usepackage{tcolorbox}
  \usepackage{colortbl}
  \usepackage{booktabs}
  \usepackage{tabularx}
  \usepackage{fontspec}
  \setmainfont [Path = fonts/,
    UprightFont = *-300,
    ItalicFont = *-300-Italic,
    BoldFont = *-700,
    BoldItalicFont = *-700-Italic
  ]{MuseoSans}
bibliography: library.bib
---

@@

Here, synthetic data plays a crucial role in asserting minimum performance requirements for novel tools in anticipation of adequate real data. Generators of scRNA-seq data (e.g. splatter[@zappia_splattersimulationsinglecell_2017], powsimR[@vieth_powsimrpoweranalysis_2017], PROSSTT[@papadopoulos_prossttprobabilisticsimulation_2018] and SymSim[@zhang_simulatingmultiplefaceted_2019]) have already been widely used to explore the strengths and weaknesses of computational tools, both by method developers[@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2018; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifyhierarchicalclassification_2019] and independent benchmarkers[@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
However, a limitation of scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities or experimental conditions.

@@

We developed a multi-modality simulator of single cells called dyngen (Figure\ \ref{fig:showcase}A). dyngen uses Gillespie's stochastic simulation algorithm[@gillespie_exactstochasticsimulation_1977]  to simulate gene regulatory interactions at a single-molecule level. Its methodology allows tracking of many layers of information throughout the simulation, including the abundance of any molecule in the cell, progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions. dyngen can simulate a large variety of dynamic processes (e.g. cyclic, branching, disconnected) as well as a broad range of experimental conditions (e.g. batch effects and time-series, perturbation and knockdown experiments). The fine-grained controls over simulation parameters allow dyngen to be applicable to a broad range of use-cases. We demonstrate this by performing first quantitative evaluations of three types of novel computational approaches: RNA velocity, casewise network inference and trajectory alignment methods.

@@

Together, these qualities allow it to be applicable in benchmarking a broad range of use-cases. To demonstrate, we apply dyngen on several novel computational approaches which have recently had a major impact on how single-cell analyses are performed but for which quantitative assessment of the performance was hitherto lacking.

## Example use cases {#sec:dyngen-example}

### Trajectory alignment

### Casewise network inference

@@

### RNA velocity

>>>>

---
title: "dyngen: benchmarking with in silico single cells"
author:
- Robrecht Cannoodt
- Wouter Saelens
- Louise Deconinck
- Yvan Saeys
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    citation_package: biblatex
biblatexoptions: sorting=none
classoption:
- table
- 10pt
- a4paper
header-includes: |
  \usepackage{tcolorbox}
  \usepackage{colortbl}
  \usepackage{booktabs}
  \usepackage{tabularx}
  \usepackage{fontspec}
  \usepackage{pifont}
  \setmainfont [Path = fonts/,
    UprightFont = *-300,
    ItalicFont = *-300-Italic,
    BoldFont = *-700,
    BoldItalicFont = *-700-Italic
  ]{MuseoSans}
bibliography: library.bib
---

@@

Here, synthetic data plays a crucial role in asserting minimum performance requirements for novel tools in anticipation of adequate real data. Generators of scRNA-seq data (e.g. splatter[@zappia_splattersimulationsinglecell_2017], powsimR[@vieth_powsimrpoweranalysis_2017], PROSSTT[@papadopoulos_prossttprobabilisticsimulation_2018] and SymSim[@zhang_simulatingmultiplefaceted_2019]) have already been widely used to explore the strengths and weaknesses of computational tools, both by method developers[@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2018; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifyhierarchicalclassification_2019] and independent benchmarkers[@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
However, a limitation of scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities or experimental conditions, thereby limiting the applicability of each generator.

```{r generators, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
modalities <- tribble(
  ~id, ~label,
  "premrna", "pre-mRNA expression",
  "mrna", "mRNA expression",
  "prot", "Protein expression",
  "promotor", "Promotor activity",
  "lineage", "Position in lineage tree",
  "reaction", "Reaction frequency"
) %>% mutate(type = "Modality")
applications <- tribble(
  ~id, ~label,
  "cluster", "Clustering",
  "diffexp", "Differential expression",
  "batch", "Batch effect correction",
  "linti", "Linear trajectory inference",
  "brati", "Branching trajectory inference",
  "disti", "Disconnected trajectory inference",
  "traja", "Trajectory alignment",
  "cni", "Casewise network inference",
  "rnav", "RNA velocity"
) %>% mutate(type = "Applications")
labels <- bind_rows(modalities, applications)

methods <- tribble(
  ~package, ~citation, ~modality, ~application, ~model,
  "splatter", "zappia_splattersimulationsinglecell_2017",
  c("mrna"), c("cluster", "linti", "batch", "diffexp"),
  "Several distributional models with parameters that allow the generation of datasets by sampling from distributions
Cell types, linear trajectories and batch effects can be simulated by changing the parameters of certain genes",

  "powsimR", "vieth_powsimrpoweranalysis_2017",
  c("mrna"), c("diffexp", "cluster"),
  "Negative binomial or zero-inflated negative binomial. Cell types can be simulated by changing the parameters of this distribution.",

  "PROSSTT", "papadopoulos_prossttprobabilisticsimulation_2018",
  c("mrna"), c("linti", "brati"),
  "(1) Generation of a random tree. (2) The expression of each gene is seen as a sum of contributions from different expression programs. The activity of each program depends on pseudotime. (3) Generation of cells and expression by random sampling along the trajectory, and from a negative binomial distribution.",

  "SymSim", "zhang_simulatingmultiplefaceted_2019",
  c("mrna", "promotor"), c("cluster", "linti", "brati", "disti"),
  "Kinetic model of each gene, in which it can be on or off, and where the mRNA can be degraded. The parameters of this model depend on a gene effect and cell effect, each of these effects represents a vector in an n-dimensional space (e.g. 10). The cell effect vectors are sampled in such a way that cells that have to be more similar (according to a tree provided by the user), will also have a more similar vector.",

  "dyngen", "",
  c("premrna", "mrna", "prot", "promotor", "lineage", "reaction"),
  c("cluster", "linti", "brati", "disti", "batch", "diffexp", "cni", "rnav", "traja"),
  "Simulation of a thermodynamic model of gene regulation, in which a set of transcription factors can bind the promoter and can have varying effects on the activity of the promoter. Clusters or trajectories are simulated by specifying a regulatory network that will induce a certain diversification of expression. Expression is simulated using a single-molecule model, after which individual UMIs are simulated using a multinomial distribution. Apart from gene regulation, splicing and protein production are also simulated."
) %>%
  mutate(
citation = ifelse(citation == "", "", paste0("\\cite{", citation, "}")),
plabel = paste0(package, " ", citation)
  )

tab <-
  full_join(
labels %>% crossing(methods) %>% select(-modality, -application),
methods %>% rowwise() %>% mutate(id = list(c(modality, application))) %>% ungroup() %>% unnest(id) %>% transmute(package, id, check = "\\ding{51}"),
by = c("id", "package")
  ) %>%
  mutate(
check = ifelse(is.na(check), "", check),
package = factor(package, levels = methods$package),
plabel = factor(plabel, levels = methods$plabel),
id = factor(id, labels$id)
  ) %>%
  arrange(match(package, methods$package), type, match(id, labels$id))

library(kableExtra)
tab %>%
  select(-id, -type, -model, -citation, -package) %>%
  spread(plabel, check) %>%
  knitr::kable(escape = FALSE) %>%
  kableExtra::row_spec(0, angle = 0)
```

@@

We developed a multi-modality simulator of single cells called dyngen (Figure\ \ref{fig:showcase}A). dyngen uses Gillespie's stochastic simulation algorithm[@gillespie_exactstochasticsimulation_1977]  to simulate gene regulation, splicing and translation at a single-molecule level. Its methodology allows tracking of many layers of information throughout the simulation, including the abundance of any molecule in the cell, progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions. dyngen can simulate a large variety of dynamic processes (e.g. cyclic, branching, disconnected) as well as a broad range of experimental conditions (e.g. batch effects and time-series, perturbation and knockdown experiments). The fine-grained controls over simulation parameters allow dyngen to be applicable to a broad range of use-cases. We demonstrate this by performing first quantitative evaluations of three types of novel computational approaches: RNA velocity, casewise network inference and trajectory alignment methods.

@@

Together, these qualities allow it to be applicable in benchmarking a broad range of use-cases. In practice, dyngen has already successfully been used to evaluate trajectory inference[@saelens_comparisonsinglecelltrajectory_2019], trajectory-based differential expression[@vandenberge_trajectorybaseddifferentialexpression_2019], and network inference[@pratapa_benchmarkingalgorithmsgene_2019] methods.
To demonstrate this point even further, we apply dyngen on several promising novel computational approaches which have recently had a major impact on how single-cell analyses are performed but for which quantitative assessment of the performance was hitherto lacking.

## Trajectory alignment

Trajectory alignment is a technique used to discover the similarities and differences between the developmental trajectories of cells. 
This technique has been used to compare gene expression kinetics resulting from different biological processes[@cacchiarelli_aligningsinglecelldevelopmental_2018], to compare human, chimpanzee and macaque neuronal development[@kanton_organoidsinglecellgenomic_2019], to find differences in gene regulation in the presence of certain growth factors[@mcfaline-figueroa_pooledsinglecellgenetic_2019], and to compare human and mouse embryogenesis[@alpert_alignmentsinglecelltrajectories_2018].

Unfortunately, not many methods to align trajectories exist. DTW (Dynamic Time Warping)[@giorgino_computingvisualizingdynamic_2009] is most commonly used to align linear trajectories. DTW is a technique originating in the field of speech recognition and aligns temporal sequences by creating a warping path between two sequences that indicates which sequence must be dilated or contracted to best match the other one.

dyngen can be used to evaluate trajectory alignment techniques. We generated two datasets and permuted each of those datasets in five ways: we knocked down the transcription factors belonging to the end of the second milestone, we set 10% or 20% of the count matrix to zero, and we performed the combination of those techniques. This results in 12 datasets, each containing a slightly different trajectory.
We compared DTW when using smoothed pseudocells (with a gaussian kernel) and using just all cells. For each combination of trajectories, we calculated the DTW alignment. In figure x, the ratio between one trajectory with a certain permutation, and the other trajectory with the same permutation is shown. We expect this ratio to be low, as we want the alignment to be able to distinguish between different trajectories. We especially expect the smoothed versions to be able to overcome the extra 10% or 20% noise.
We performed this experiment 10 times.

In Figure \ref{fig:traj_align}, we can see that DTW with gaussian smoothing cannot distinguish between trajectory 1 and trajectory 2. When using the cells themselves, we find that they can clearly distinguish between trajectory 1 and 2 when noise is added, but not when the trajectory is limited to milestone C.

<!--
\begin{figure}[htb!]
    \centering
    \includegraphics[width=.8\linewidth]{fig/traj_alignment/tares.pdf}
    \caption{
    \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
    \label{fig:traj_align}
\end{figure}
-->

## Casewise network inference

@@

## RNA velocity
